Reasoning About a Machine with Local Capabilities

### 2019-03-06 toplas/paper.tex
Add an ML-version of the awkward example and use it to explain the assembly version.

The final example, \texttt{g1} with \texttt{f4}, is a faithful translation of a tricky example known from the literature (known as the awkward example)~\citep{pitts_operational_1998,Dreyer:jfp12}.
It consists of two parts, \texttt{g1} and \texttt{f4}.
The program \texttt{g1} is a closure generator that generates closures with one variable $x$ set to $0$ in its environment and \texttt{f4} as the program (note we omit some calling convention security measures because the stack is not used in the closure generator).
The program\texttt{f4} expects one argument, a callback.
\texttt{f4} sets $x$ to $0$ and calls the callback.
When the callback returns, \texttt{f4} sets $x$ to $1$ and calls the callback again.
When the callback returns again, \texttt{f4} asserts $x$ is $1$ and returns.
This example is more complicated than the previous ones because it involves a closure invoked by the adversary and an adversary callback invoked by us.
\itoplassug{ Suggestion: Maybe give some more intuition as to why it is difficult.
} As explained in \sectionname~\ref{sec:stack-and-return-pointer}, we need to check that (1) the stack pointer that the closure receives from the adversary has write-local permission and (2) the adversary callback is global.

To illustrate how subtle this program is, consider how an adversary could try to
make the assertion fail. In the second callback an adversary can get to the
first callback by invoking the closure one more time. If there were any way for
the adversary to transfer the return pointer from the point where it reinvokes
the closure to where the closure reinvokes the callback, then the assertion
could be made to fail. Similarly, if there were any way for the adversary to
store a stack pointer or trick the trusted code into preserving it across an
invocation, the assertion could be made to fail too. However, our calling
convention prevents any of this from happening, as we prove in the following
lemma.

>>>>

The final example, \texttt{g1} with \texttt{f4}, is a faithful translation of a tricky example known from the literature (known as the awkward example)~\citep{pitts_operational_1998,Dreyer:jfp12}.
For comparison, we show a version of the original example in Figure~\ref{fig:awkward}, highlighting the code locations corresponding to \texttt{g1} and \texttt{f4}.

\begin{wrapfigure}{R}{.35\textwidth}
  \centering
  \begin{tabular}{c} % see https://tex.stackexchange.com/questions/5818/how-to-center-a-listing
  \begin{lstlisting}[gobble=2,escapeinside=||]
  fun _ => 
  |\color{gray}{g1:}| let x = ref 0 in
      fun adv =>
  |\color{gray}{f4:}|   x := 0;
        adv();
        x := 1;
        adv();
        assert(x == 1)
  \end{lstlisting}
  \end{tabular}
  \caption{The original awkward example from \citet{pitts_operational_1998,Dreyer:jfp12}, in ML notation.}
  \label{fig:awkward}
\end{wrapfigure}

Let us first look at this ML program.
At the top-level, it is a lambda function that can be invoked by the context.
When it is invoked, it allocates a fresh mutable variable $x$ of integer type, initially containing the value $0$.
Next, the function returns a second closure (let's call this $\textit{cl}_x$) that can be invoked by the context whenever it chooses.
$\textit{cl}_x$ itself takes a callback function \texttt{adv} that it will invoke twice, after setting $x$ to $0$ and $1$ respectively.
After the second invocation of \texttt{adv}, $\textit{cl}_x$ will verify that $x$ is still set to $1$.

This last assertion, at line 8 of Figure~\ref{fig:awkward}, should never fail.
From the code, this seems natural, since \texttt{adv} does not have access to $x$.
Therefore, after the second invocation at line 7, $x$ should still be in the state it was before that invocation.
However, consider that \texttt{adv} may have access to $\textit{cl}_x$ and within that second invocation of \texttt{adv}, it may invoke $\textit{cl}_x}$ again.
Then $\textit{cl}_x$ would set $x$ to $0$ again (on line 5) and reinvoke another callback.
If that second callback were somehow able to skip its own caller and return directly to the caller of \texttt{adv}, it would end up at line 8 of Figure~\ref{fig:awkward} with $x$ set to $1$, causing the assertion to fail.
Without going into details, such an attack is perfectly possible if \texttt{adv} has access to a call/cc primitive (or equivalent).
In other words, if we are able to prove that the assertion will never fail, this attack, and other similar attacks, on well-bracketed control flow are adequately prevented.

Our low-level version of the awkward example in Figure~\ref{fig:prog-f3-and-g1}, consists of two parts, \texttt{g1} and \texttt{f4}, corresponding to the code locations marked in Figure~\ref{fig:awkward}.
The program \texttt{g1} corresponds to the top-level closure in the ML code.
It is a closure generator that generates closures with a mutable variable $x$ set to $0$ in its environment and \texttt{f4} as the program (note that we omit some calling convention security measures because the stack is not used in the closure generator).

The program \texttt{f4} expects one argument, the callback \texttt{adv}, in \texttt{r1}.
\texttt{f4} sets $x$ to $0$ and invokes \texttt{adv}.
When it returns, \texttt{f4} sets $x$ to $1$ and calls \texttt{adv} again.
When it returns the second time, \texttt{f4} asserts $x$ is still $1$ and returns.

This example is more complicated than the previous ones because it involves a closure invoked by the adversary and an adversary callback invoked by us.
As explained in \sectionname~\ref{sec:stack-and-return-pointer}, we need to check that (1) the stack pointer that the closure receives from the adversary has write-local permission and (2) the adversary callback is global.

The attack that we explained above is actually more natural at this low-level.
The callback \texttt{adv} now gets a return capability.
As explained, it could invoke $\mathit{cl}_x$ again, during its second invocation, with a second callback function $\texttt{adv}'$ for $\mathit{cl}_x$ to invoke.
If \texttt{adv} had some way of passing its return capability to $\texttt{adv}'$ (by storing it somewhere on the stack or in the heap, or hide it an unused register), then the assertion could be made to fail.
However, our calling convention prevents any of this from happening, as we prove in the following lemma.
